Assessing Phonon Properties of Shape Memory Alloys using Graph Neural Network Potentials

### 2023-12-29 src/tex/ms.tex
New scripts, potential, rework Snakemake, move folders, added ms text.

%% Packages
\usepackage[utf8]{inputenc}
\usepackage{fontspec}
\usepackage{amsmath}
\usepackage{siunitx}
\usepackage[version=4]{mhchem}
\usepackage{pdflscape}
%% Handles multipage table. Not sure allowed by pub.
\usepackage{longtable}

@@

% Abstract with filler text
\begin{abstract}
Shape Memory Alloys (SMAs) such as NiTi are pivotal in vibrational dampening,robotic arms, prothestic hands, and mechanical vales as a result of their unique displacive phase transformation properties. However, assessing their phonon characteristics, which provide insight into the displacive phase transformation, by using conventional Density Functional Theory (DFT) methods is computationally challenging. In this paper we assess the use of recent Graph Neural Network (GNN) potentials as a viable alternative. These GNN potentials have leveraged large DFT-computed databases for training and provide a good approximation to the potential energy surfaces which enable inference on energies and forces in lattice and atomic dynamic calculations. We compare GNN potentials with the Embedded Atom Method (EAM) in predicting phonon properties of NiTi. We find that in general the EAM potentials are slightly more performant than the GN in that they appear to more accurately describe equation of state, phonon dispersions, and elastic constants. However, the MACE potential shows significant promise as an exploritory potential for designing and characterizing more complex SMAs. Our findings reveal GNN potentials as a promising avenue for advancing SMA research and fine-tuning of pre-trained GNN models with more relevant DFT data (i.e., force-constants) may permit better outcomes.  
\end{abstract}

@@

Recently, there as been a growing effort in the development of machine learning potentials. These deviate from traditional physics approaches to interatomic potentials, which have historically been inspired from more accurate methods like tight-binding method. In machine learning potentials, one describes the atoms and structure using descriptors of features, these are typically learned or prescribed.

The main inquiry of this work is to asses the GNN interatomic potentials on binary SMA NiTi and PbTi primarily through the phonon predictions. Our analysis aims to provide a qualitative perspective based on previously studied phonon phenomena of these systems. The findings here will provide guidnace for future use of GNN interatomic potentials for the development other binary, and more so, ternary and quaternary SMAs. IMportantly, the use of pretrained base models for these systems acts as an initial seed and can be used to fine-tune future GNN potentials for better SMA disocovery works. \par

@@

\subsection{Workflow}
\label{sec:workflow}
The ground-state structures for the NiTi phases \textit{B2}, \textit{B19}, \textit{B19'}, and \textit{BCO} are generated using the atomic simulation environment (ASE) package (cite). This requires that the unit cell parameters, basis atoms, and spacegroup number be provided. These initial reference parameters for these structures have been taken from the materials project (cite specific DOIs for the structures). The energy and forces on the atoms are determined using the calculator interface routines which enables the assesment of different interatomic potentials for the same ASE structure. Similarly, for all EOS, phonon, and elastic constant claculations each step is performed for a specific interatomic potential via the ASE calculator interface. We not that this is done in different python computing environments using \href{https://github.com/showyourwork/showyourwork}{\showyourwork} package.\par

First the structures are optimized using ASE local optimization routines. The forces were converged to $10^{-4}\,\text{eV}/\AA$ and utlize the FIRE optimizater, but found other local optimization options found same ground-state structures. To ensure that the unit cell parameters were adjusted along with the atomic coordinates and that the crystal symmetry was not broken, we utlized the ASE constraint and filter routines during optimization of the structures. \par

To determine the equation of state, we started with the obtained ground-state structures. To adjust the volume of the unit cell, we then apply an isotropic strain tensor to the cell and scale the atomic coordinates

\begin{equation}
  \label{eq:isotropic_strain}
  T=\begin{bmatrix}
    \epsilon & 0 & 0 \\
    0 & \epsilon & 0 \\
    0 & 0 & \epsilon \\
  \end{bmatrix}
\end{equation}

were $\epsilon$ is the fraction strain applied to the unit cell. To ensure that the strain is indeed isotropic we provide a quality check on the symmetry of the distorted structure. The isotropic strain is applied in compression and tension, from $-11\%$ to $11\%$, which gave a good range to asses the EOS curves and obtain fits. We fit the EOS to a standard polynomial which provides values for the equilibrium volume, energy, and bulk modulus. In order to compare respective equations of state amongst the different phases, we normalize the volume of each phase with respect to the obtained ground-state volume. \par

@@

It is observed that for all but the ALIGNN potential, that upon high tensile isotropic strain, the \textit{Pbcm} phase is the most stable. \par

@@

\begin{figure}[ht!]
    \script{PlotEOS.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/NiTi_EquationOfStates.png}
        \caption{
            The equation of state curves for different \ce{NiTi} structures for (a) Mutter (EAM) (b) Zhong (EAM) (c) M3GNet (d) CHGNet (e) MACE and (f) ALIGNN interatomic potentials. The x-axis is the ratio of the isotropic strained unit cell volume, $V$,  with respect to the ground-state unit cell volume, $V_o$.
        }
        \label{fig:eos}
    \end{centering}
\end{figure}

@@

\begin{figure}[ht!]
    \script{PlotPhonons.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/NiTi_B2_ModelsPhononBandstructures.png}
        \caption{
          Ground-state, no isotropic strain, Phonon bandstructure of B2 phase for all models. The Mutter and Zhong EAM potentials differ in that Mutter predicts no dynamic instability of the B2 phase at 0K. 
        }
        \label{fig:allmodels_b2}
    \end{centering}
\end{figure}

\begin{figure}[ht!]
    \script{PlotPhonons.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/NiTi_B19_ModelsPhononBandstructures.png}
        \caption{
          Ground-state phonon bandstructure of B19 phase for all models.
        }
        \label{fig:allmodels_B19P}
    \end{centering}
\end{figure}

\begin{figure}[ht!]
    \script{PlotPhonons.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/NiTi_B19P_ModelsPhononBandstructures.png}
        \caption{
          Ground-state phonon bandstructure of \textit{B19}$^\prime$ phase for all models.
        }
        \label{fig:allmodels_B19P}
    \end{centering}
\end{figure}

\begin{figure}[ht!]
    \script{PlotPhonons.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/NiTi_BCO_ModelsPhononBandstructures.png}
        \caption{
          Ground-state phonon bandstructure of BCO phase for all models.
        }
        \label{fig:allmodels_B19P}
    \end{centering}
\end{figure}

\begin{figure}[ht!]
    \script{PlotEAMPhonons.py}
    \begin{centering}
        \includegraphics[width=0.5\linewidth]{figures/NiTi_Mutter_B2_StrainsPhononBandstructures.png}
        \vspace{1mm}
        \includegraphics[width=0.5\linewidth]{figures/NiTi_Zhong_B2_StrainsPhononBandstructures.png}
        \caption{
           Phonon dispersion of B2 structure for the Mutter (top) and Zhong (bottom)  EAM potentials at different isotropic strains. The optical branches have been made more transparent to emphasis the behavior in the TA$_1$ and TA$_2$ branches. The color indicates direction from compressive to tensile strain. Note that the strain percent shown is with respect to isotropic changes (i.e. volume). 
        }
        \label{fig:mutter_zhong_phonon_b2}
    \end{centering}
\end{figure}

@@

\begin{figure}[ht!]
    \script{PlotGNNPhonons.py}
    \begin{centering}
      \includegraphics[width=0.5\linewidth]{figures/NiTi_M3GNet_B2_StrainsPhononBandstructures.png}
      \vspace{1mm}
      \includegraphics[width=0.5\linewidth]{figures/NiTi_CHGNet_B2_StrainsPhononBandstructures.png}
      \vspace{1mm}
      \includegraphics[width=0.5\linewidth]{figures/NiTi_MACE_B2_StrainsPhononBandstructures.png}
      \caption{
        Phonon dispersion of B2 structure for the (a) M3GNet, (b) CHGNet, and (c) MACE potentials for different isotropic strains. Similar transcparency is used as done in \ref{fig:mutter_zhong_phonon_b2}.
      }
      \label{fig:gnn_phonon_b2}
    \end{centering}
\end{figure}

@@

\subsection{PbTi: Equation of States}

@@

\begin{figure}[ht!]                                                                                      
    \script{AppendixBZ.py}                                                                           
    \begin{centering}                                                                                    
      \includegraphics[width=0.5\linewidth]{figures/B2_BrillouinZonePointsSampled.png} 
      \vspace{1mm}                                                                                       
      \includegraphics[width=0.5\linewidth]{figures/B19_BrillouinZonePointsSampled.png}
      \vspace{1mm}
      \includegraphics[width=0.5\linewidth]{figures/BCO_BrillouinZonePointsSampled.png}
      \vspace{1mm}                                                                                       
      \caption{The irreducible Brillouin zones for (top) B2(\#221) (middle) B19/B19P(\#11), and (bottom) BCO(\#63) crystal structures}                                                                                  \label{fig:ibz}                                                                          
    \end{centering}                                                                                      
\end{figure}

\variable{output/B2_SpecialSymmetryPointsBZ.tex}
\variable{output/B19_SpecialSymmetryPointsBZ.tex}
\variable{output/BCO_SpecialSymmetryPointsBZ.tex}

>>>>

%% Packages
\usepackage[utf8]{inputenc}
\usepackage{fontspec}
\usepackage{amsmath}
\usepackage{siunitx}
\usepackage[version=4]{mhchem}
\usepackage{pdflscape}
%% Handles multipage table. Not sure allowed by pub.
\usepackage{longtable}
\usepackage{subcaption}

@@

% Abstract with filler text
\begin{abstract}
Shape Memory Alloys (SMAs) such as NiTi are pivotal in vibrational dampening,robotic arms, prothestic hands, and mechanical vales as a result of their unique displacive phase transformation properties. However, assessing their phonon characteristics, which provide insight into the displacive phase transformation, by using conventional Density Functional Theory (DFT) methods is computationally challenging. In this paper we assess the use of recent Graph Neural Network (GNN) potentials as a viable alternative. These GNN potentials have leveraged large DFT-computed databases for training and provide a good approximation to the potential energy surfaces which enable inference on energies and forces in lattice and atomic dynamic calculations. We compare GNN potentials with the Embedded Atom Method (EAM) and Modified EAM style potentials in predicting phonon properties of NiTi and PtTi. We find that in general the EAM/MEAM potentials are slightly more performant than the GNN in that they appear to more accurately describe equation of state, phonon dispersions, and elastic constants. However, the MACE potential shows significant promise as an exploritory potential for designing and characterizing more complex SMAs. Our findings suggest GNN potentials as a promising initial avenue for advancing SMA research, in particular ternary and quaternary systems, upon fine-tuning of pre-trained GNN models with more relevant DFT data (i.e., force-constants) to permit better outcomes.   
\end{abstract}

@@

Recently, there as been a growing effort in the development of machine learning potentials. These deviate from traditional physics approaches to interatomic potentials, which have historically been inspired from more accurate methods like tight-binding method. In machine learning potentials, one describes the atoms and structure using descriptors of features, these are typically learned or prescribed. The features can be prescribed in isolation or combination of  structural, compositional, and property characteristics of the material data point. Futhermore, one may choose to fined new representations via unsurpervised learning. \par

The subsequent step once the featureization process is completed is to determine a suitable neural network archetcture which can learn the target property or properties given the feature inputs. In the case of atomistic simulations the target is prediction of the potential energy surface and corresponding gradients (i.e., forces). Two of the most early attempts to do this were the works of \citeauthor{Bartok2010} and \citeauthor{Thompson2015}, whereby projection onto higher-dimensional basis, such as bispectrum componenets of local neighbor density, is learned in order to describe the atomic environments. The energy per atom is then determined by these learned features by employing linear regression or non-linear Gaussian process regression. \par

While these early attempts were successfully, they could only include structural details to learn new features from. Futhermore, the training process was typically restricted to a limited number of elemental constieunts and therefore for each chemical system and phases of interest one needs to provide a suitable training dataset. To move beyond such limitations, groups began using graph neural network (GNN) architectures and message passing scheme to represent the atomic structures as nodes and edges. This enabled two distinctions, first being that the nodes and edges corresponding to the atoms and bonds could include features, and secondly the learning could be done over a large structural and compositional space enabling a generalization\footnote{Typically these are referred to as Universal interatomic potentials or forcefields since they in theory generalize well to any chemical system or phase.}. The use of GNNs is efficient in representing structure-composition-property inputs to additional NN layers for predicting potential energy surfaces and gradients and was initially pursued in the CGCNN, MEGNet, and ALIGNN potentials.\cite{Choudhary2021,Chen2019,Xie2018}. The differences between the approaches are predominately in atomic and bond features, nearest-neighbor graph, and specific NN archetcture. \par

The training of these early GNN utlized large density functional theory calculation repositories and mostly tuned the learning cost function to ground-state structure and energy. This resulted in mean-absolute-errors and other measures of performance that could be improved upon.\cite{Riebesell2023} More recently there have been significant improvements in both training data used and the GNN architecture that show promise for truly approaching a general interatomic potential suitable for static and dynamic atomistic calculations of multicomponents and multiphases. If succesfull, identifying and designing new potential SMA materials using GNN interatomic potentials could have tremendous impact on technology. Therefore, the primary goal is to provide a qualitative assement of predictions of the equations of state, phonon bandstructure, and elastic constants for various phases of NiTi and PtTi using three GNN potentials intended for general use: M3GNet, CHGNet, and MACE.\cite{Chen2022,Deng2023,Batatia2022} We note that the \cite{Riebesell2023} provides a summarized performance over these GNN over materials project database, but does not specifically address SMA and phonon characteristics. \par

@@

\subsection{Workflow}
\label{sec:workflow}
The ground-state structures for the NiTi phases \textit{B2}, \textit{B19}, \textit{B19'}, and \textit{BCO} are generated using the atomic simulation environment (ASE) package \cite{Larsen2017}. This requires that the unit cell parameters, basis atoms, and spacegroup number be provided. The initial unit cell parameters and basis corrdinates for these structures have been taken from the materials project\cite{Jain2013} and several references \cite{Haskins2016,Kadkhodaei2018}. The initial structures for NiTi are shown in fig~\ref{fig:niti_structures} and are adapted for PtTi as well given same spacegroup symmetries. The energy and forces on the atoms are determined using the ASE calculator interfaces for EAM, MEAM, M3GNet, CHGNet, and MACE.\cite{Mutter2010,Zhong2011,Ko2015,Kim2017,Chen2022,Deng2023,Batatia2022} The EAM and MEAM potential interfaces are achieved through the use of LAMMPS python library\cite{Thompson2022} and ASE. The ALIGNN ASE calculator interface was also implemented, however, we found that the predictions for NiTi and PtTi to be unreliable.\par

For each combination of potential/model, chemical system, and phase we perform calcualtions for the structure minimization, equation of state (EOS), phonons, and elastic constants. To determine the finite-temperature phase transitions we utlize an isothermal-isobaric ensemble via molecular dynamics simulations. The thermostating and barostating is done with Nose-Hoover and Parrinello-Rahman dynamics and the tempertaure is adjusted in a quasi-static manner which entails equlibrating the simulation for 10 ns while decrementing the temperature by 10 Kelvin; the pressure is kept constant at 1 bar. We note that in order to achieve the different computing environments required by the GNN potentials we leverage the ability to specify the  python computing environments while using \href{https://github.com/showyourwork/showyourwork}{\showyourwork} package.\par

\begin{figure}[ht!]
    \script{VisualizedStructures.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/NiTi_VisualizedStructures.png}
        \caption{
          Initial unit cells of used in this work for NiTi, (a) \textit{B2}(221), (b) \textit{B19}(11), (c) \textit{B19'}(51), and (d) \textit{BCO}(61). The spacegroup numbers are given in the parentheses. Also additional atoms (i.e. non-basis sites) are shown based on periodic boundary conditions. For PtTi (a-b) structures are used with different unit cell parameters and basis coordinates. 
        }
        \label{fig:niti_structures}
    \end{centering}
\end{figure}

Prior to any property calculations or dynamics the structures are optimized using the FIRE\cite{Bitzek2006} local optimizer with the forces converged to $2\cdot 10^{-3}\,\text{eV}/\AA$. We observed that other local optimization routines found same ground-state structures. To ensure that the unit cell parameters were adjusted along with the atomic coordinates and that the crystal symmetry was not broken, we utlized a constraint and filter during the optimization process. \par

The equation of state of state can be obtained by varying the cell volume and calculating the total system energy without allowing for atomic relaxations. From the obtained ground-state structures the unit cell volume is adjusted by applying an isotropic strain,

\begin{equation}
  \label{eq:isotropic_strain}
  T=\begin{bmatrix}
    \epsilon & 0 & 0 \\
    0 & \epsilon & 0 \\
    0 & 0 & \epsilon \\
  \end{bmatrix}
\end{equation},

to the cell and scaling the fractional atomic coordinates. Here $\epsilon$ is the fraction strain applied to the unit cell. To ensure that the strain is indeed isotropic we provide a quality check on the symmetry of the distorted structure using the spacegroup package \textit{spglib}.\cite{Togo2018} The isotropic strain is applied in compression and tension, from $-11\%$ to $11\%$, which gave a good range to asses the EOS curves and obtain fits. We fit the EOS to a third order inverse polynomial,

\begin{equation}
  \label{eq:eos_fit}
  E(V) = c_0 + c_1 t + c_2 t^2 + c_3 t^3, \quad \text{where} \quad t = V^{-\frac{1}{3}}
\end{equation},

where the fit coefficients are used to determine equilibrium volume, system energy, and bulk modulus \cite{Alchagirov2003}. In order to compare respective equations of state amongst the different phases, we normalize the strained volume of each phase with respect to the obtained minimized ground-state volume. \par

@@

@@

\begin{figure}[ht!]
    \script{PlotEOS.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/NiTi_EquationOfStates.png}
        \caption{
            The equation of state curves for different \ce{NiTi} structures for (a) Mutter (EAM) (b) Zhong (EAM) (c) Ko (2NN-MEAM) (d) M3GNet (e) CHGNet (f) MACE interatomic potentials. The x-axis is the ratio of the isotropic strained unit cell volume, $V$,  with respect to the ground-state unit cell volume, $V_o$.
        }
        \label{fig:eos}
    \end{centering}
\end{figure}

@@

\begin{figure}[!htp]
    \script{PlotPhonons.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/NiTi_B2_ModelsPhononBandstructures.png}
        \caption{
          Ground-state, no isotropic strain, Phonon bandstructure of B2 phase for all models. The Mutter and Zhong EAM potentials differ in that Mutter predicts no dynamic instability of the B2 phase at 0K. 
        }
        \label{fig:allmodels_b2}
    \end{centering}
\end{figure}

\begin{figure}[!htp]
    \script{PlotPhonons.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/NiTi_B19_ModelsPhononBandstructures.png}
        \caption{
          Ground-state phonon bandstructure of B19 phase for all models.
        }
        \label{fig:allmodels_B19P}
    \end{centering}
\end{figure}

\begin{figure}[!htp]
    \script{PlotPhonons.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/NiTi_B19P_ModelsPhononBandstructures.png}
        \caption{
          Ground-state phonon bandstructure of \textit{B19}$^\prime$ phase for all models.
        }
        \label{fig:allmodels_B19P}
    \end{centering}
\end{figure}

\begin{figure}[!htp]
    \script{PlotPhonons.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/NiTi_BCO_ModelsPhononBandstructures.png}
        \caption{
          Ground-state phonon bandstructure of BCO phase for all models.
        }
        \label{fig:allmodels_B19P}
    \end{centering}
\end{figure}

\begin{figure}[!htp]
    \script{PlotPhonons.py}
    \begin{centering}
        \includegraphics[width=0.5\linewidth]{figures/NiTi_Mutter_B2_StrainsPhononBandstructures.png}
        \vspace{1mm}
        \includegraphics[width=0.5\linewidth]{figures/NiTi_Zhong_B2_StrainsPhononBandstructures.png}
        \vspace{1mm}
        \includegraphics[width=0.5\linewidth]{figures/NiTi_Ko_B2_StrainsPhononBandstructures.png}
        \caption{
           Phonon dispersion of B2 structure for the  (top) EAM Mutter, (middle) EAM Zhong, and (bottom) 2NN-MEAM Ko potentials at different isotropic strains. The optical branches have been made more transparent to emphasis the behavior in the TA$_1$ and TA$_2$ branches. The color indicates direction from compressive to tensile strain. Note that the strain percent shown is with respect to isotropic changes (i.e. volume). 
        }
        \label{fig:mutter_zhong_phonon_b2}
    \end{centering}
\end{figure}

@@

\begin{figure}[!htp]
    \script{PlotPhonons.py}
    \begin{centering}
      \includegraphics[width=0.5\linewidth]{figures/NiTi_M3GNet_B2_StrainsPhononBandstructures.png}
      \vspace{1mm}
      \includegraphics[width=0.5\linewidth]{figures/NiTi_CHGNet_B2_StrainsPhononBandstructures.png}
      \vspace{1mm}
      \includegraphics[width=0.5\linewidth]{figures/NiTi_MACE_B2_StrainsPhononBandstructures.png}
      \caption{
        Phonon dispersion of B2 structure for the (top) M3GNet, (middle) CHGNet, and (bottom) MACE potentials for different isotropic strains. Similar transcparency is used as done in \ref{fig:mutter_zhong_phonon_b2}.
      }
      \label{fig:gnn_phonon_b2}
    \end{centering}
\end{figure}

@@

\subsection{PtTi: Equation of States}

@@

\begin{figure}[!htp]                                                                                      
    \script{AppendixBZ.py}                                                                           
    \begin{centering}                                                                                    
      \includegraphics[width=0.5\linewidth]{figures/B2_BrillouinZonePointsSampled.png} 
      \vspace{1mm}                                                                                       
      \includegraphics[width=0.5\linewidth]{figures/B19_BrillouinZonePointsSampled.png}
      \vspace{1mm}
      \includegraphics[width=0.5\linewidth]{figures/B19P_BrillouinZonePointsSampled.png}
      \vspace{1mm}
      \includegraphics[width=0.5\linewidth]{figures/BCO_BrillouinZonePointsSampled.png}
      \vspace{1mm}                                                                                       
      \caption{The irreducible Brillouin zones for \textit{B2}(\#221), \textit{B19}(\#51), \textit{B19'}(\#11), and \textit{BCO}(\#63) crystal structures}                                                               \label{fig:ibz}                                                                          
    \end{centering}                                                                                      
\end{figure}

\variable{output/B2_SpecialSymmetryPointsBZ.tex}
\variable{output/B19_SpecialSymmetryPointsBZ.tex}
\variable{output/B19P_SpecialSymmetryPointsBZ.tex}
\variable{output/BCO_SpecialSymmetryPointsBZ.tex}
